---
title: "Citius, Altius, Fortius in NCAA Track & Field"
subtitle: "Are NCAA athletes getting faster, higher, & stronger?"
author: "TS O'Leary"
output:
  rmarkdown::html_document:
    theme: sandstone
    toc: false
    toc_float: false
runtime: shiny
---

```{r setup, include = FALSE}
# don't show cose, warnings or messages
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)

# Load packages
require(tidyverse)
require(shiny)
require(broom)
require(kableExtra)

shinyThings::radioSwitchButtons_default_style(selected_background = "#faa555")

# Load data
df <- readRDS(url("https://tsoleary.github.io/track/NCAA/data/tfrrs_NCAA_2010_2021_final_data_team.rds"))

# Create split data by event type -- TRACK, FIELD, MULTI -- for analysis
df_split <- df %>%
  group_by(EVENT_TYPE) %>%
  group_split()

# Colors for plotting teams
df_colors <- read_csv("https://raw.githubusercontent.com/tsoleary/track/main/NCAA/data/ncaa_colors.csv")


# Define sprint events for plotting
sprints <- unique(df$EVENT)[c(1:3, 9:10, 12)]


# Define event order for good plotting
event_order <- c("100m", 
                 "100m Hurdles",
                 "110m Hurdles",
                 "4 x 100m Relay",
                 "200m",
                 "400m",
                 "400m Hurdles",
                 "4 x 400m Relay",
                 "800m",
                 "1500m",
                 "3,000m Steeplechase",
                 "5,000m",
                 "10,000m", 
                 "High Jump",
                 "Pole Vault",
                 "Long Jump",
                 "Triple Jump",
                 "Shot Put",
                 "Discus",
                 "Hammer",
                 "Javelin",
                 "Heptathlon",
                 "Decathlon")


# Define functions -------------------------------------------------------------

# Function: violin_box_plot -----
# Description: Creates a violin box plot
# Inputs: input_description
# Outputs: output_description

# Required packages
require(tidyverse)

# Example call
# df %>%
#   filter(EVENT == "Long Jump" & SEX == "Men") %>%
#   violin_box_plot(y = "MARK_METERS", y_lab = "Meters")


violin_box_plot <- function(dat, 
                            y,
                            x_lab = "", 
                            y_lab = y) {
  ggplot(dat,
         aes_string(x = "YEAR_QUAL", 
                    y = y)) +
    geom_violin(fill = "azure2", 
                color = "grey40") +
    geom_boxplot(fill = "azure2", 
                 color = "grey40", 
                 width = 0.2) +
    labs(x = x_lab, y = y_lab,
         title = paste0(unique(dat$SEX), "'s ", unique(dat$EVENT))) +
    theme_classic(base_size = 16)
} ## End function -----


# Function: lm_events_calc -----
# Description: linear model on all events with scaled or unscaled options
# Inputs: split tibble with EVENT, SEX, MARK_METERS, TIME_S, or POINTS columns
#   n_top_ath - number of top athletes to include in the model
#   scale - logical whether or not to scale before the linear model
# Outputs: tidy lm model output

# Required packages
require(tidyverse)
require(broom)

# Example call
# lm_10_scaled <- lm_events_calc(df_split, top_n_ath = 10, scaled = TRUE)


lm_events_calc <- function(dat, top_n_ath, scaled = FALSE) {
  
  # To scale or not to scale
  if (scaled) {
  # Field events -- linear regression
  df_field_lm <- df_split[[1]] %>%
    group_by(EVENT, SEX) %>%
    mutate(MARK_METERS = scale(MARK_METERS)) %>%
    filter(RANK <= top_n_ath) %>%
    mutate(YEAR_QUAL = as.numeric(as.character(YEAR_QUAL))) %>%
    nest() %>%
    mutate(
      fit = map(data, ~ lm(MARK_METERS ~ YEAR_QUAL, data = .x)),
      tidied = map(fit, tidy)
    ) %>%
    unnest(tidied) %>%
    mutate(EVENT_TYPE = "FIELD")
 
  # Multi events -- linear regression
  df_multi_lm <- df_split[[2]] %>%
    group_by(EVENT, SEX) %>%
    mutate(POINTS = scale(POINTS)) %>%
    filter(RANK <= top_n_ath) %>%
    mutate(YEAR_QUAL = as.numeric(as.character(YEAR_QUAL))) %>%
    nest() %>%
    mutate(
      fit = map(data, ~ lm(POINTS ~ YEAR_QUAL, data = .x)),
      tidied = map(fit, tidy)
    ) %>%
    unnest(tidied) %>%
    mutate(EVENT_TYPE = "MULTI")
  
  # Track events -- linear regression
  df_track_lm <- df_split[[3]] %>%
    group_by(EVENT, SEX) %>%
    mutate(TIME_S = scale(TIME_S)) %>%
    filter(RANK <= top_n_ath) %>%
    mutate(YEAR_QUAL = as.numeric(as.character(YEAR_QUAL))) %>%
    nest() %>%
    mutate(
      fit = map(data, ~ lm(TIME_S ~ YEAR_QUAL, data = .x)),
      tidied = map(fit, tidy)
    ) %>%
    unnest(tidied) %>%
    mutate(EVENT_TYPE = "TRACK")
  
  } else {
    
  # Field events -- linear regression
  df_field_lm <- df_split[[1]] %>%
    filter(RANK <= top_n_ath) %>%
    group_by(EVENT, SEX) %>%
    mutate(YEAR_QUAL = as.numeric(as.character(YEAR_QUAL))) %>%
    nest() %>%
    mutate(
      fit = map(data, ~ lm(MARK_METERS ~ YEAR_QUAL, data = .x)),
      tidied = map(fit, tidy)
    ) %>%
    unnest(tidied) %>%
    mutate(EVENT_TYPE = "FIELD")
 
  # Multi events -- linear regression
  df_multi_lm <- df_split[[2]] %>%
    filter(RANK <= top_n_ath) %>%
    group_by(EVENT, SEX) %>%
    mutate(YEAR_QUAL = as.numeric(as.character(YEAR_QUAL))) %>%
    nest() %>%
    mutate(
      fit = map(data, ~ lm(POINTS ~ YEAR_QUAL, data = .x)),
      tidied = map(fit, tidy)
    ) %>%
    unnest(tidied) %>%
    mutate(EVENT_TYPE = "MULTI")
  
  # Track events -- linear regression
  df_track_lm <- df_split[[3]] %>%
    filter(RANK <= top_n_ath) %>%
    group_by(EVENT, SEX) %>%
    mutate(YEAR_QUAL = as.numeric(as.character(YEAR_QUAL))) %>%
    nest() %>%
    mutate(
      fit = map(data, ~ lm(TIME_S ~ YEAR_QUAL, data = .x)),
      tidied = map(fit, tidy)
    ) %>%
    unnest(tidied) %>%
    mutate(EVENT_TYPE = "TRACK")
  }
  
  # Combine all results and add info about scale and n_top_ath
  df_lm <- bind_rows(df_track_lm, df_field_lm, df_multi_lm) %>%
    mutate(scale = scaled,
           top_n_ath = as.factor(top_n_ath))
  
  # Return results
  return(df_lm)
  
} ## End function -----

```

# {.tabset}

## Graphic

```{r shiny}
shinyApp(
  
  # Define the user interface
  ui <- fluidPage(
    
    # titlePanel("Test"),
    
    sidebarLayout(
      
      sidebarPanel(width = 3,
        
        # Choose the Event
        selectInput(
          inputId = "event", 
          label = "Event",
          choices = c(unique(df$EVENT)[1:21], "Multi"), 
          selected = "800m"
          ),
        
        # Choose Sex
        shinyThings::radioSwitchButtons(
          inputId = "sex",
          choices = c("Men" = "Men", "Women" = "Women"),
          selected = "Women"
          ),

        # Choose rank
        sliderInput(
          inputId = "rank",
          label = "Top # Athletes",
          min = 0,
          max = 500,
          value = 100,
          step = 10,
          ),
        
        # Choose team(s) to add to plot
        selectInput(
          inputId = "teams", 
          label = "Teams",
          choices = unique(df$ncaa_name), 
          multiple = TRUE,
          selected = NULL
          ),
      
      # Viz trend?
      shinyWidgets::prettySwitch(
        inputId = "trend",
        label = "Visualize trendline",
        fill = TRUE, 
        status = "primary"
      ),
        
        ),
      
      # Plot  
      mainPanel(width = 9,
                plotOutput(outputId = "plot"))
    )
  ),
  
  
  # Define the server
  server <- function(input, output, session) {
    
    
    # Filter data based on inputs
    data <- reactive({
      if (input$event == "Multi") {
        if (input$sex == "Men") {
          temp <- df %>% 
            filter(EVENT == "Decathlon" & SEX == input$sex) %>% 
            filter(RANK <= input$rank)
        } else {
          temp <- df %>% 
            filter(EVENT == "Heptathlon" & SEX == input$sex) %>% 
            filter(RANK <= input$rank)
        }
      } else {
          temp <- df %>% 
            filter(EVENT == input$event & SEX == input$sex) %>% 
            filter(RANK <= input$rank)
      }
      
      temp

      })
    
    # Filter data based on inputs
    data_team <- reactive({
      dat <- data()
      
      if (length(input$teams > 0)) {
        temp <- dat %>%
          filter(ncaa_name %in% input$teams) %>%
          group_by(ncaa_name) %>%
          group_split()
      } else {
        temp <- NULL
      }
      
      temp

      })
    
    # Plot
    output$plot <- renderPlot({
      temp <- data() 
      
      if (unique(temp$EVENT_TYPE) == "TRACK") {
        if (unique(temp$EVENT) %in% sprints) {
          p <- temp %>%
            violin_box_plot(y = "TIME_S", 
                            y_lab = "Time (sec)")  + 
            scale_y_continuous(labels = scales::label_number(accuracy = 0.01))
          if (input$trend == TRUE) {
            p <- p +
              geom_smooth(data = temp, 
                          aes(x = as.numeric(YEAR_QUAL), 
                              y = TIME_S),
                          method = lm,
                          se = FALSE,
                          formula = y ~ x,
                          color = "grey50")
          }
        } else {
          p <- temp %>%
            violin_box_plot(y = "TIME_S", 
                            y_lab = "Time (min:sec)") +
            scale_y_time(labels = function(l) strftime(l, '%M:%S'))
          if (input$trend == TRUE) {
            p <- p +
              geom_smooth(data = temp, 
                          aes(x = as.numeric(YEAR_QUAL), 
                              y = TIME_S),
                          method = lm,
                          se = FALSE,
                          formula = y ~ x,
                          color = "grey50")
          }
        }

      } else if (unique(temp$EVENT_TYPE) == "FIELD") {
          p <- temp %>%
            violin_box_plot(y = "MARK_METERS", 
                            y_lab = "Mark (m)") + 
            scale_y_continuous(labels = scales::label_number(accuracy = 0.01))
          if (input$trend == TRUE) {
            p <- p +
              geom_smooth(data = temp, 
                          aes(x = as.numeric(YEAR_QUAL), 
                              y = MARK_METERS),
                          method = lm,
                          se = FALSE,
                          formula = y ~ x,
                          color = "grey50")
          }
        
      } else {
          p <- temp %>%
            violin_box_plot(y = "POINTS", 
                            y_lab = "Points")
         if (input$trend == TRUE) {
            p <- p +
              geom_smooth(data = temp, 
                          aes(x = as.numeric(YEAR_QUAL), 
                              y = POINTS),
                          method = lm,
                          se = FALSE,
                          formula = y ~ x,
                          color = "grey50")
          }
        
      }
      
      # Define teams
      dat <- data_team()
      
      if (length(dat) > 0) {
        
        # Define colors for plotting
        cols <- df_colors %>%
          filter(ncaa_name %in% input$teams)
        
      if (unique(temp$EVENT_TYPE) == "TRACK") {
        if (unique(temp$EVENT) %in% sprints) {
          for (i in 1:length(dat)) {
            p <- p +
               geom_jitter(data = dat[[i]], 
                           width = 0.1,
                           aes(x = YEAR_QUAL, 
                               y = TIME_S),
                           alpha = 0.9,
                           size = 2.5, 
                           shape = 21, 
                           stroke = 1,
                           fill = as.character(cols[cols$ncaa_name == 
                                         unique(dat[[i]]$ncaa_name),
                                       "primary_color"]),
                           color = as.character(cols[cols$ncaa_name == 
                                          unique(dat[[i]]$ncaa_name),
                                        "secondary_color"]))
          }
        } else {
          for (i in 1:length(dat)) {
            p <- p +
               geom_jitter(data = dat[[i]], 
                           width = 0.1,
                           aes(x = YEAR_QUAL, 
                               y = TIME_S),
                           alpha = 0.9,
                           size = 2.5, 
                           shape = 21, 
                           stroke = 1,
                           fill = as.character(cols[cols$ncaa_name == 
                                         unique(dat[[i]]$ncaa_name),
                                       "primary_color"]),
                           color = as.character(cols[cols$ncaa_name == 
                                         unique(dat[[i]]$ncaa_name),
                                       "secondary_color"]))
          }
        }

      } else if (unique(temp$EVENT_TYPE) == "FIELD") {
          for (i in 1:length(dat)) {
            p <- p +
               geom_jitter(data = dat[[i]], 
                           width = 0.1,
                           aes(x = YEAR_QUAL, 
                               y = MARK_METERS),
                           alpha = 0.9,
                           size = 2.5, 
                           shape = 21, 
                           stroke = 1,
                           fill = as.character(cols[cols$ncaa_name == 
                                         unique(dat[[i]]$ncaa_name),
                                       "primary_color"]),
                           color = as.character(cols[cols$ncaa_name == 
                                          unique(dat[[i]]$ncaa_name),
                                        "secondary_color"]))
          }
        
      } else {
          for (i in 1:length(input$teams)) {
            p <- p +
               geom_jitter(data = dat[[i]], 
                           width = 0.1,
                           aes(x = YEAR_QUAL, 
                               y = POINTS),
                           alpha = 0.9,
                           size = 2.5, 
                           shape = 21, 
                           stroke = 1,
                           fill = as.character(cols[cols$ncaa_name == 
                                         unique(dat[[i]]$ncaa_name),
                                       "primary_color"]),
                           color = as.character(cols[cols$ncaa_name == 
                                          unique(dat[[i]]$ncaa_name),
                                        "secondary_color"]))
          }
        }
      }
    # Print the plot
    p
   })
  },

  # Set height for output of panel in html
  options = list(height = 450)
)
```

## About

### Data & code

All data was collected from [TFRRS](https://www.tfrrs.org/). Data represent the best performance of the top 500 NCAA DI athletes in each event at the time of qualifying for the First Round (Regionals) of the NCAA Outdoor Track & Field Championship -- in other words, these data do not include results from the NCAA Championship. All code used for this analysis is open-source and available on [this GitHub repository](https://github.com/tsoleary/track) -- please contact me if you are interested in collaborating.

### Questions I have of these data

- Which events are improving at the fastest rate?
- Are the best athletes getting better at a faster rate than the rest of the top athletes?
- Are women improving faster than men?
- How does class year (year of eligibility) affect performance?
  - To what degree is an athlete's eligibility year predictive of their rank?
  - Is eligibility year more predictive for certain events?
  - Do some events skew younger or older?
  - Does the effect of eligibility year differ between men and women?
- Which individual teams are improving in rank faster than others?
- Are the Power Five Conferences increasing in rank faster than other conferences?


# 

```{r}
# Scaled results
lm_10_scaled <- lm_events_calc(df_split, top_n_ath = 10, scaled = TRUE)
lm_100_scaled <- lm_events_calc(df_split, top_n_ath = 100, scaled = TRUE)
lm_500_scaled <- lm_events_calc(df_split, top_n_ath = 500, scaled = TRUE)

# Raw results
lm_10_raw <- lm_events_calc(df_split, top_n_ath = 10, scaled = FALSE)
lm_100_raw <- lm_events_calc(df_split, top_n_ath = 100, scaled = FALSE)
lm_500_raw <- lm_events_calc(df_split, top_n_ath = 500, scaled = FALSE)

# Combine results
res <- bind_rows(lm_10_scaled, lm_100_scaled, lm_500_scaled,
                 lm_10_raw, lm_10_raw, lm_10_raw)
```


```{r, eval = FALSE}
# RESULTS SUMMARIES ----
x <- res %>% 
  filter(term == "YEAR_QUAL") %>% 
  filter(p.value < 0.05/252) %>%
  arrange(p.value)

x %>%
  select(EVENT, SEX, scale, top_n_ath, estimate, p.value) %>%
  mutate(p.value = as.character(signif(p.value, 2))) %>%
  kable() %>%
  kable_styling("striped") %>% 
  scroll_box(width = "100%", height = "250px") 

x %>%
  group_by(SEX, top_n_ath) %>%
  count() %>%
  pivot_wider(names_from = top_n_ath, values_from = n) %>%
  kable() %>%
  kable_styling("striped") %>%
  add_header_above(c(" ", "Top # Athletes" = 3))

x %>%
  group_by(EVENT_TYPE, top_n_ath) %>%
  count() %>%
  pivot_wider(names_from = top_n_ath, values_from = n) %>%
  kable() %>%
  kable_styling("striped") %>%
  add_header_above(c(" ", "Top # Athletes" = 3))

x %>%
  group_by(EVENT_TYPE, SEX) %>%
  count() %>%
  pivot_wider(names_from = SEX, values_from = n) %>%
  kable() %>%
  kable_styling("striped")

```

# Average change per year

## {.tabset}

### Women

```{r, fig.height = 7}
res %>%
  filter(SEX == "Women" & 
           scale == TRUE & 
           term == "YEAR_QUAL" &
           p.value < 0.05/252) %>%
  mutate(EVENT = factor(EVENT, levels = rev(event_order))) %>%
  ggplot(aes(x = EVENT, 
             y = estimate,
             ymin = 0,
             ymax = estimate,
             color = top_n_ath)) + 
  geom_point(position = position_dodge(width = 0.5),
             stat = 'identity', 
             size = 2) +
  geom_linerange(position = position_dodge(width = 0.5)) +
  scale_color_manual(name = "Top #\nAthletes",
                     values = c("grey10", "grey30", "grey60")) +
  #scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(y = "Standard change in mean performance per year",
       x = element_blank()) +
  coord_flip() +
  theme_minimal() 
```

### Men

```{r, fig.height = 7}
res %>%
  filter(SEX == "Men" & 
           scale == TRUE & 
           term == "YEAR_QUAL" &
           p.value < 0.05/252) %>%
  mutate(EVENT = factor(EVENT, levels = rev(event_order))) %>%
  ggplot(aes(x = EVENT, 
             y = estimate,
             ymin = 0,
             ymax = estimate,
             color = top_n_ath)) + 
  geom_point(position = position_dodge(width = 0.5),
             stat = 'identity', 
             size = 2) +
  geom_linerange(position = position_dodge(width = 0.5)) +
  scale_color_manual(name = "Top #\nAthletes",
                     values = c("grey10", "grey30", "grey60")) +
  labs(y = "Standard change in mean performance per year",
       x = element_blank()) +
  coord_flip() +
  theme_minimal()
```

## Men vs. Women  {.tabset}

### Top 10

```{r, fig.height = 7}
res %>%
  filter(scale == TRUE &
           top_n_ath == 10 &
           term == "YEAR_QUAL" &
           p.value < 0.05/252) %>%
  mutate(EVENT = factor(EVENT, levels = rev(event_order))) %>%
  ggplot(aes(x = EVENT, 
             y = estimate,
             ymin = 0,
             ymax = estimate,
             color = SEX)) + 
  geom_point(position = position_dodge(width = 0.5),
             stat = 'identity', 
             size = 2) +
  geom_linerange(position = position_dodge(width = 0.5)) +
  scale_color_manual(name = "Sex",
                     values = c("darkseagreen3", "mediumpurple1")) +
  labs(y = "Standard change in mean performance per year",
       x = element_blank()) +
  coord_flip() +
  theme_minimal()
```

### Top 100

```{r, fig.height = 7}
res %>%
  filter(scale == TRUE &
           top_n_ath == 100 &
           term == "YEAR_QUAL" &
           p.value < 0.05/252) %>%
  mutate(EVENT = factor(EVENT, levels = rev(event_order))) %>%
  ggplot(aes(x = EVENT, 
             y = estimate,
             ymin = 0,
             ymax = estimate,
             color = SEX)) + 
  geom_point(position = position_dodge(width = 0.5),
             stat = 'identity', 
             size = 2) +
  geom_linerange(position = position_dodge(width = 0.5)) +
  scale_color_manual(name = "Sex",
                     values = c("darkseagreen3", "mediumpurple1")) +
  labs(y = "Standard change in mean performance per year",
       x = element_blank()) +
  coord_flip() +
  theme_minimal()
```

### Top 500

```{r, fig.height = 7}
res %>%
  filter(scale == TRUE &
           top_n_ath == 500 &
           term == "YEAR_QUAL" &
           p.value < 0.05/252) %>%
  mutate(EVENT = factor(EVENT, levels = rev(event_order))) %>%
  ggplot(aes(x = EVENT, 
             y = estimate,
             ymin = 0,
             ymax = estimate,
             color = SEX)) + 
  geom_point(position = position_dodge(width = 0.5),
             stat = 'identity', 
             size = 2) +
  geom_linerange(position = position_dodge(width = 0.5)) +
  scale_color_manual(name = "Sex",
                     values = c("darkseagreen3", "mediumpurple1")) +
  labs(y = "Standard change in mean performance per year",
       x = element_blank()) +
  coord_flip() +
  theme_minimal()
```

